When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-01-24 doc/paper/vpd_et_paper_PG_v1.tex
at one interesting sid enote

The ET reponse to changes in VPD alters water partitioning between the soil and atmosphere. If plant response reduces ET in response to atmospheric drying then soil moisture will be better conserved. This would seem a sensible evolutionary strategy to cope with aridity. If stomata were fully passive \citep [similar to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would further reduce soil moisture. In turn, this would further increase aridity as low soil moisture levels increase the Bowen ratio, and in turn increase temperature and reduce atmospheric humidity (gentine et al. GRL 2016, Berg 2016).  This however would not seem to be a sensible strategy for plants from an evolutionary standpoint.

@@

Differences in the sign term between plant types is exclusively determined by $uWUE$ and the slope parameter ($g_1$). A larger $uWUE$ means that at the ecosystem scale $WUE$ is larger for a given VPD, so plants are more willing to keep stomata open for new growth despite high VPD. Similiarly, high slope parameters ($g_1$) means that at the leaf scale plants are more willing to trade water for growth. Both these effects (large $uWUE$, $g_1$) conspire to shift the sign term towards overall positive values for a given PFT.

This effect of $uWUE$ and $g1$ on the sign term is most apparent by comparing CRO to CSH. CRO has a high slope parameter ($g_1 = 183.1 \, Pa^{1/2}$) and uWUE ($2.32$ -units-) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$, $uWUE=2.11$ -units-). These differences in slope parameter cause opposite ET reponses to changse in VPD. ET almost always increases with increasing VPD for CRO, while ET almost always decreases with increasing VPD for CSH (Figure \ref{term3}). CRO, evolved or bred to prioritize GPP, are very willing to trade water for new growth, despite changes in VPD, while CSH are very unwilling to trade water for new growth.

The slope paramter ($g_1$) is additionally a primary determinant of VPD  dependnece for the sign term shown in Figure \ref{term3}. Plants that are less willing

Based on Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the most positive constant offset, while CSH are the most water conservative and have the most negative constant offset in the sign term.  ENF ($g_1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g_1 = 183.1$) has the smallest VPD dependence. ENF are less willing to trade water for new growth, so the regulate water much more strongly depending on the dryness environment (VPD). CRO on the otherhand are more willing to trade water for new growth, so they regulate water relatively similiarly irregardless of the environment.

A primary takeaway from Figure \ref{term3}a is that according to our theory for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. This thus means that plants are able in most atmospheric conditions to redue ET in response to increased VPD and thus to reduce water losses. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates and empathizes the water conservative strategy of those plants. For CRO on the other, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on the environment, with about half of of observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for potential evaporation i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface reponse to a change in VPD will always me more negative than if the plants are not included in the physics (PET). However, we have assumed $\sigma$ (uncertainty) equal to one. We still need to test if inclusion of uncertainty could change our conclusions.
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***need to replot with correct sigma (normalized to 1), also correct uWUE label in first plot***}}
\label{term3}
\end{figure}

Figure \ref{term3}b shows the location of VPD$_{crit}$, as a function of $\sigma$ (uncertainty) and $VPD$. For any $\sigma$ or VPD less (more) than these curves, the sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. This motivates a thorough examination of how inclusion of uncertainty ($\sigma$) might change the form of the sign term, and in particular the location of VPD$_{crit}$ for each of our plant types. In Secion \ref{testing} we examine if and how uncertainty changes the analysis presented in this section, for each PFT.

@@

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; VPD}$ will also the relative magnitude of the change $\frac{\partial \; ET}{\partial \; VPD}$ and the importance of $VPD$ variability for overall $ET$ variability. So we now more closely examine the scaling term: $\frac{P}{T} \propto \rho$. Air density $\rho$ scaling term varies little relative to aerodynamic conductance and $\Delta$. The psychrometric constant $\gamma$ is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the slope of the Clauisu-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency fo the transfer of surface anomalies to the atmosphere. As aerodynamic conductace  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

For DBF our theory is successful towards this goal: the theory matches the leading order behavior of the function when uncertainty is included, and the DBG figure looks as one would expect if one just added noise on top of our theory. The VPD dependence, given by the slope parameter ($g_1$), floows the median values of each box plot. Perhaps most importantly, the x-intercept, and thus VPD$_{crit}$, matches nearly exactly between the theory and the observations. Therefor the sign of the ET response to increases in VPD should be well matched, subject to the unavoidable constraints of noise, much of which may be from the observations themselves. The uncertainty is non negligable; there are bany observations in each box for which the the sign of observation is opposite the response predicted by the theory, but to leading order our theory matches the observations.

@@

The comparison between the theory and obsercations for ENF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope paramter, is not well captured by observations. However, the theory closely captures the x-intercept and VPD$_{crit}$. So, the theory still captures when the ET reponse will be positive or negative, but is less accurate with the magnitude of the reponse. Therefor the theory is still useful for addressing when VPD drives or reduces ET. Also, it's with noting that we used published constants for $g_1$ \citep[from ][]{Lin_2015}, which determines the VPD dependence. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori physical constants whenever possible.

>>>>

The ET response to changes in VPD alters water partitioning between the soil and atmosphere. If plant response reduces ET in response to atmospheric drying then soil moisture will be better conserved. This would seem a sensible evolutionary strategy to cope with aridity. If stomata were fully passive \citep [similar to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would further reduce soil moisture. In turn, this would further increase aridity as low soil moisture levels increase the Bowen ratio, and in turn increase temperature and reduce atmospheric humidity (gentine et al. GRL 2016, Berg 2016).  This however would not seem to be a sensible strategy for plants from an evolutionary standpoint.

@@

Differences in the sign term between plant types is exclusively determined by $uWUE$ and the slope parameter ($g_1$). A larger $uWUE$ causes means that at the ecosystem scale $WUE$ decreases more with increasing VPD, so plants lose more water per unit carbon production and are more likely to increase ET with increasing VPD. A tendenacy towards increasing ET reponse with increasing VPD can also be caused by high slope parameters ($g_1$), characteristic of plants that at the leaf scale plants are more willing to trade water for growth. Plants that give water more freely will be much more likely to increase ET with increasing VPD.  Both the aforementioned effects (large $uWUE$, $g_1$) can amplify each other, and generally conspire to shift the sign term towards overall positive values for a given PFT.

This effect of $uWUE$ and $g1$ on the sign term is most apparent by comparing CRO to CSH. CRO has a high slope parameter ($g_1 = 183.1 \, Pa^{1/2}$) and uWUE ($2.32$ -units-) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$, $uWUE=2.11$ -units-). These differences in slope parameter cause opposite ET reponses to changse in VPD between CRO and CSH. ET almost always increases with increasing VPD for CRO, while ET almost always decreases with increasing VPD for CSH (Figure \ref{term3}). CRO, evolved or bred to prioritize GPP, are very willing to trade water for new growth, despite changes in VPD, while CSH are very unwilling to trade water for new growth.

The slope paramter ($g_1$) is additionally a primary determinant of VPD  dependnece for the sign term shown in Figure \ref{term3}. Plants that are less willing to trade water for new growth (small $g_1$) will have strong tendency to reduce ET with increasing VPD, and they will be very effective in reducing ET especially at low VPD.  However, at very high VPD gradients in vapor pressure at the leaf scale will begin to dominate leaf-scale plant strategies (parameterized with $g_1$), so ET reponse will begin to asymtote towards ecosystem-scale values irregardless of leaf-scale strategy (and $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependece of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{term3}.

To summarize Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the strongest overall tendency to increase ET wirh increasing VPD, while CSH are the most water conservative and have the strongest tendency to decrease ET with increasing VPD.  ENF ($g_1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g_1 = 183.1$) has the smallest VPD dependence.  ENF are less willing to trade water for new growth at the leaf scale, so the difference between the response at low VPD when leaf response domineates and high VPD when vapor pressure gradients dominate is much greater. CRO on the otherhand are more willing to trade water for new growth, so the leaf scale response is much smaller, and the difference in ET response between low VPD and high VPD is relatively low.

Figure \ref{term3} show clearly that, accoridng to our thoery, for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. Therefore, plants are able in most atmospheric conditions to redue ET in response to increased VPD and thus to reduce water loss. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates and empathizes the water conservative strategy of those plants. For CRO on the otherhand, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on the environment, with about half of of observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for potential evaporation i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface reponse to a change in VPD will always me more negative than if the plants are not included in the physics (PET). 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***I think I'm going to remove the bottom plot: its discussion didn't fit in well with the text, and the box plots test inclusion of uncertainty***}}
\label{term3}
\end{figure}

@@

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, understaing the overall magnitude of the ET reponse is important to the soil-plant-atmosphere water budgeting. So we now more closely examine the terms that affect how the sign term is scaled:

$\frac{P}{T}$ is an air-density term, which we would expect to vary relatively little compared to aerodynamic conductance and $\Delta$. The psychrometric constant ($\gamma$) is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the slope of the Clauisu-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency fo the transfer of the surface environment to the atmosphere. As aerodynamic conductace  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

For DBF our theory is successful towards this goal: the theory matches the leading order behavior of the function when uncertainty is included, and the DBG figure looks as one would expect if one just added noise on top of our theory. The VPD dependence, given by the slope parameter ($g_1$), floows the median values of each box plot. Perhaps most importantly, the x-intercept, and thus VPD$_{crit}$, matches nearly exactly between the theory and the observations. Therefore the sign of the ET response to increases in VPD should be well matched, subject to the unavoidable constraints of noise, much of which may be from the observations themselves. The uncertainty is non negligable; there are bany observations in each box for which the the sign of observation is opposite the response predicted by the theory, but to leading order our theory matches the observations.

@@

The comparison between the theory and obsercations for ENF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope paramter, is not well captured by observations. However, the theory closely captures the x-intercept and VPD$_{crit}$. So, the theory still captures when the ET reponse will be positive or negative, but is less accurate with the magnitude of the reponse. Therefore the theory is still useful for addressing when VPD drives or reduces ET. Also, it's with noting that we used published constants for $g_1$ \citep[from ][]{Lin_2015}, which determines the VPD dependence. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori physical constants whenever possible.
